---
title: "obsplot Gallery : marks"
author: "Julien Barnier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 4.5
    toc: true
vignette: >
  %\VignetteIndexEntry{obsplot Gallery : marks}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
library(obsplot)
library(htmlwidgets)
knitr::opts_chunk$set(
  screenshot.force = FALSE,
  echo = TRUE
)
```

This document is an attempt to reproduce some of the charts shown in [Observable Plot documentation notebooks](https://observablehq.com/@observablehq/plot).

As always, there are several ways to do things, in particular data manipulation could be done either in R before plotting, or in JavaScript via transforms and `JS()` calls. Each example below is one way to achieve a given result, not necessarily the best one or the most elegant.

## Area mark

```{r}
data(aapl)
obsplot(aapl) |>
    mark_areaY(x = "Date", y = "Close", fill = "#ccc") |>
    mark_lineY(x = "Date", y = "Close") |>
    mark_ruleY(y = 0) |>
    scale_y(grid = TRUE)
```


```{r}
data(sftemp)

obsplot(sftemp) |>
    mark_ruleY(y = 32) |>
    mark_areaY(
        transform_windowY(
            x = "date",
            y1 = "low", y2 = "high",
            fill = "#ccc", k = 14
        )
    ) |>
    mark_line(
        transform_windowY(
            x = "date",
            y = JS("d => (d.low + d.high) / 2"),
            k = 14
        )
    ) |>
    scale_y(grid = TRUE, label = "↑ Temperature (°F)")

```


```{r}
data(bls)

obsplot(bls) |>
    mark_areaY(x = "date", y2 = "unemployed", z = "industry", fillOpacity = 0.1) |>
    mark_lineY(x = "date", y = "unemployed", z = "industry", strokeWidth = 1)
```


```{r}
obsplot(bls) |>
    mark_areaY(x = "date", y = "unemployed", fill = "industry") |>
    mark_ruleY(y = 0)
```

```{r}
normY <- transform_normalizeY(basis = "median", x = "date", y = "unemployed")
obsplot(bls, height = 600) |>
    mark_areaY(normY, fillOpacity = 0.2) |>
    mark_lineY(normY, strokeWidth = 1) |>
    facet(y = "industry", marginRight = 140) |>
    scale_y(axis = NULL) |>
    scale_fy(axis = "right", label = NULL)
```


## Bar mark


```{r}
data(civilizations)

obsplot(civilizations, height = nrow(civilizations) * 12) |>
    mark_barX(x1 = "start", x2 = "end", y = "civilization") |>
    mark_text(
        x = "start", y = "civilization", text = "civilization",
        textAnchor = "end", dx = -6
    ) |>
    opts(marginLeft = 140) |>
    scale_x(axis = "top", grid = TRUE) |>
    scale_y(axis = NULL, domain = civilizations$civilization[order(civilizations$start)])
```

```{r}
data(popchange)
popchange$diff <- (popchange$`2019` - popchange$`2010`) / popchange$`2010` * 100
popchange$up <- popchange$diff >= 0

obsplot(popchange, height = 780) |>
    mark_barX(y = "State", x = "diff", fill = "up") |>
    mark_ruleX(x = 0) |>
    opts(marginLeft = 110, grid = TRUE) |>
    scale_x(
        axis = "top", round = TRUE,
        label = "← decrease · Change in population, 2010–2019 (%) · increase →",
        labelAnchor = "center", tickFormat = "+"
    ) |>
    scale_y(label = NULL, domain = popchange$State[order(popchange$diff)], reverse = TRUE) |>
    scale_color(range = c("#e15759", "#4e79a7"))

```

```{r}
data(alphabet)

obsplot(alphabet, height = 60) |>
    mark_ruleX(c(0,1)) |>
    mark_barX(
        transform_stackX(order = "letter", x = "frequency", fill = "#ccc", insetLeft = 1)
    ) |>
    mark_textX(
        transform_stackX(order = "letter", x = "frequency", text = "letter", insetLeft = 1)
    ) |>
    scale_x(label = "Frequency (%)", transform = JS("d => d * 100"))

```


```{r message=FALSE}
data(stateage)

# Precompute state names order
library(dplyr)
states <- stateage |>
    group_by(name) |>
    summarize(total = sum(population)) |>
    arrange(desc(total)) |>
    slice(1:6) |>
    pull(name)

obsplot(stateage) |>
    mark_barY(x = "age", y = "population", fill = "age", title = "age") |>
    mark_ruleY(y = 0) |>
    facet(x = "name") |>
    scale_x(axis = NULL, domain = unique(stateage$age)) |>
    scale_fx(domain = states, label = NULL, tickSize = 6) |>
    scale_y(grid = TRUE, tickFormat = "s") |>
    scale_color(domain = unique(stateage$age), scheme = "spectral")
```


## Cell mark


```{r}
obsplot(simpsons, height = 600) |>
    mark_cell(x = "season", y = "number_in_season", fill = "imdb_rating") |>
    mark_text(x = "season", y = "number_in_season", text = "imdb_rating", title = "title") |>
    scale_x(axis = "top", label = "Season") |>
    scale_y(label = "Episode") |>
    scale_color(scheme = "PiYg") |>
    opts(padding = 0.05, grid = TRUE)
```

```{r message = FALSE}
data(dji)
# precompute variables
dji <- dji |>
    mutate(
        variation = (Close - lag(Close)) / lag(Close),
        title = round(variation * 100, 1)
    )

obsplot(dji, height = 1300) |>
    mark_cell(
        x = JS("d => d3.utcWeek.count(d3.utcYear(d.Date), d.Date)"),
        y = JS("d => d.Date.getUTCDay()"),
        fill = "variation",
        title = "title",
        inset = 0.5
    ) |>
    facet(y = JS("d => d.Date.getUTCFullYear()")) |>
    scale_x(axis = NULL, padding = 0) |>
    scale_y(
        padding = 0, tickSize = 0,
        tickFormat = JS('Plot.formatWeekday("en", "narrow")')
    ) |>
    scale_fy(reverse = TRUE, label = NULL) |>
    scale_color(type = "diverging", scheme = "PiYg")
```

```{r}
data(seattle)

obsplot(seattle, height = 300) |>
    mark_cell(
        transform_group(
            list(fill = "max"),
            x = JS("d => d.date.getUTCDate()"),
            y = JS("d => d.date.getUTCMonth()"),
            fill = "temp_max",
            inset = 0.5
        )
    ) |>
    scale_y(
        tickFormat = JS('Plot.formatMonth("en", "short")')
    ) |>
    opts(padding = 0)
```

```{r}
df <- simpsons |>
    mutate(tick = season * 100 + number_in_season - 1)

obsplot(df, height = 60) |>
    mark_cell(x = "tick", fill = "imdb_rating") |>
    scale_x(
        ticks = df |> filter(number_in_season == 1) |> pull(tick),
        tickFormat = JS("d => d / 100"),
        round = FALSE, label = "Season →", labelAnchor = "right"
    ) |>
    scale_color(scheme = "PiYg")
```


## Dot mark


```{r}
data(gistemp)

obsplot(gistemp) |>
    mark_ruleY(y = 0) |>
    mark_dot(x = "Date", y = "Anomaly", stroke = "Anomaly") |>
    scale_y(tickFormat = "+f", grid = TRUE) |>
    scale_color(type = "diverging", scheme = "BuRd")
```


```{r}
data(aapl)

obsplot(aapl) |>
    mark_ruleX(0) |>
    mark_dot(
        x = JS("d => (d.Close - d.Open) / d.Open"),
        y = "Volume", r = "Volume"
    ) |>
    scale_x(
        label = "Daily change (%) →",
        tickFormat = "+f", transform = JS("d => d * 100")
    ) |>
    scale_y(
        label = "↑ Daily trading volume",
        type = "log", tickFormat = "~s"
    ) |>
    opts(grid = TRUE)
```

```{r}
data(diamonds_obs)

obsplot(diamonds_obs, height = 480) |>
    mark_dot(
        transform_bin(r = "count", x = "carat", y = "price", thresholds = 100)
    ) |>
    scale_x(label = "Carats →") |>
    scale_y(label = "↑ Price ($)") |>
    scale_r(range = c(0, 20)) |>
    opts(grid = TRUE)
```

```{r}
data(driving)
# filter out highlighted data beforehand
driv5 <- driving |> filter(year %% 5 == 0)

obsplot(driving) |>
    mark_line(x = "miles", y = "gas", curve = "catmull-rom") |>
    mark_dot(x = "miles", y = "gas", fill = "currentColor") |>
    mark_text(driv5, x = "miles", y = "gas", text = "year", dy = -8) |>
    opts(grid = TRUE)
```


```{r}
data(alphabet)

obsplot(alphabet) |>
    mark_ruleY(y = 0) |>
    mark_ruleX(x = "letter", y = "frequency") |>
    mark_dot(x = "letter", y = "frequency", fill = "black", r = 4) |>
    scale_x(label = NULL, tickSize = 0) |>
    scale_y(transform = JS("d => d * 100"), label = "↑ Frequency (%)")
```


```{r}
data(stateage)

xy <- transform_normalizeY(basis = "sum", z = "name", x = "age", y = "population")
obsplot(stateage) |>
    mark_ruleY(y = 0) |>
    mark_line(xy, strokeWidth = 1, stroke = "#ccc") |>
    mark_dot(xy) |>
    scale_x(domain = unique(stateage$age), labelAnchor = "right") |>
    scale_y(transform = JS("d => d*100")) |>
    opts(grid = TRUE)
```

```{r}
xy <- transform_normalizeX(basis = "sum", z = "name", x = "population", y = "name")

obsplot(stateage, height = 660) |>
    mark_ruleX(x = 0) |>
    mark_ruleY(
        transform_groupY(list(x1 = "min", x2 = "max"), xy)
    ) |>
    mark_dot(xy, fill = "age", title = "age") |>
    mark_text(
        transform_selectMinX(xy), textAnchor = "end", dx = -6, text = "name"
    ) |>
    scale_x(axis = "top",   label =  "Percent (%) →", transform = JS("d => d * 100")) |>
    scale_y(axis = NULL) |>
    scale_color(scheme = "spectral", domain = unique(stateage$age)) |>
    opts(grid = TRUE)
```

## Link mark

```{r}
data(metros)
# Compute the difference first
metros$diff1580 <- metros$R90_10_2015 - metros$R90_10_1980

obsplot(metros) |>
    mark_link(
        x1 = "POP_1980", y1 = "R90_10_1980",
        x2 = "POP_2015", y2 = "R90_10_2015", stroke = "diff1580"
    ) |>
    mark_dot(x = "POP_2015", y = "R90_10_2015", r = 1) |>
    mark_text(
        x = "POP_2015", y = "R90_10_2015",
        filter = "highlight", text = "nyt_display", dy = -6
    ) |>
    scale_x(type = "log", tickFormat = "~s") |>
    scale_color(type = "diverging", reverse = TRUE) |>
    opts(grid = TRUE)
```

```{r}
data(income)
# Compute min and max values first
imin <- min(c(income$m, income$f), na.rm = TRUE)
imax <- max(c(income$m, income$f), na.rm = TRUE)
qs <- c(.6, .7, .8, .9, 1)

obsplot(height = 600) |>
    mark_link(x1 = imin, y1 = imin, x2 = imax, y2 = imax) |>
    mark_link(
        x1 = imin, y1 = imin * qs, x2 = imax, y2 = imax * qs,
        strokeOpacity = ifelse(qs == 1, 1, 0.2)
    ) |>
    mark_text(
        x = imax, y = qs * imax, textAnchor = "start", dx = 6,
        text = ifelse(qs == 1, "Equal", qs)
    ) |>
    mark_dot(income, x = "m", y = "f", title = paste(income$type, income$age)) |>
    opts(marginRight = 40) |>
    scale_x(
        label = "Median annual income (men, thousands) →",
        tickFormat = JS("d => d / 1000")
    ) |>
    scale_y(
        label = "↑ Median annual income (women, thousands)",
        tickFormat = JS("d => d / 1000")
    )
```

## Line mark

```{r}
data(aapl)

obsplot(aapl) |>
    mark_line(x = "Date", y = "Close") |>
    scale_y(grid = TRUE)
```

```{r}
obsplot() |>
    mark_lineY(JS("d3.cumsum({length: 300}, d3.randomNormal())")) |>
    scale_x(axis = NULL)
```

```{r}
data(bls_unemployment)

obsplot(bls_unemployment) |>
    mark_ruleY(y = 0) |>
    mark_line(x = "date", y = "unemployment", z = "division") |>
    scale_y(grid = TRUE, label = "↑ Unemployment (%)")
```

```{r}
data(stocks)

obsplot(stocks) |>
    mark_ruleY(1) |>
    mark_line(
        transform_normalizeY(x = "Date", y = "Close", stroke = "Symbol")
    ) |>
    mark_text(
        transform_selectLast(
            transform_normalizeY(
                x = "Date", y = "Close", stroke = "Symbol",
                text = "Symbol", textAnchor = "start", dx = 3
            )
        )
    ) |>
    scale_y(
        type = "log", grid = TRUE, label = "↑ Change in price (%)",
        tickFormat = JS("x => d3.format('+d')((x - 1) * 100)")
    ) |>
    style(overflow = "visible")
```

```{r}
data(bls_unemployment)

obsplot(bls_unemployment) |>
    mark_ruleY(0) |>
    mark_line(
        x = "date", y = "unemployment", z = "division", stroke = "unemployment"
    ) |>
    scale_y(grid = TRUE, label = "↑ Unemployment (%)")
```

```{r}
data(bls_unemployment)
bls_unemployment$highlight <- grepl(", MI ", bls_unemployment$division)

obsplot(bls_unemployment) |>
    mark_ruleY(0) |>
    mark_line(
        x = "date", y = "unemployment", z = "division",
        stroke = "highlight", sort = "highlight"
    ) |>
    scale_y(grid = TRUE, label = "↑ Unemployment (%)") |>
    scale_color(domain = c(FALSE, TRUE), range = c("#CCC", "red"))
```

```{r}
data(sftemp)

obsplot(sftemp) |>
    mark_line(
        transform_windowY(k = 14, x = "date", y = "low", stroke = "#4e79a7")
    ) |>
    mark_line(
        transform_windowY(k = 14, x = "date", y = "high", stroke = "#e15759")
    ) |>
    mark_ruleY(32) |>
    scale_y(grid = TRUE, label = "↑ Temperature (°F)")
```

## Rect mark

```{r}
data(aapl)

obsplot(aapl) |>
    mark_rectY(
        transform_binX(
            y = "count", x = JS("d => Math.log10(d.Volume)"), normalize =  TRUE
        )
    ) |>
    mark_ruleY(0) |>
    scale_x(round = TRUE, label = "Trade volume (log₁₀) →") |>
    scale_y(grid = TRUE)
```


```{r}
data(povcalnet)

obsplot(povcalnet) |>
    mark_rectY(
        transform_stackX(
            filter = JS("d => ['N', 'A'].includes(d.CoverageType)"),
            x = "ReqYearPopulation",
            order = "HeadCount",
            reverse = TRUE,
            y2 = "HeadCount",
            title = JS("d => `${d.CountryName}\n${(d.HeadCount * 100).toFixed(1)}%`"),
            insetLeft = 0.2,
            insetRight = 0.2
        )
    ) |>
    scale_x(label = "Population (millions) →") |>
    scale_y(
        label = "↑ Proportion living on less than $30 per day (%)",
        transform = JS("d => d * 100"),
        grid = TRUE
    )
```


```{r}
data(diamonds_obs)

obsplot(diamonds_obs, height = 640) |>
    mark_rect(
        transform_bin(
            fill = "count", x = "carat", y = "price", thresholds = 100, inset = 0
        )
    ) |>
    opts(round = TRUE)  |>
    scale_color(scheme = "bupu", type = "symlog")
```

```{r}
data(aapl)

obsplot(aapl) |>
    mark_rectY(
        transform_binX(
            y = "count", rx = 8, x = JS("d => Math.log10(d.Volume)"), normalize =  TRUE
        )
    ) |>
    mark_ruleY(0) |>
    scale_x(round = TRUE, label = "Trade volume (log₁₀) →") |>
    scale_y(grid = TRUE)
```

## Rule mark

```{r}
obsplot(height = 60) |>
    mark_ruleX(data = list(length = 500), x = JS("d3.randomNormal()"), strokeOpacity = 0.2) |>
    scale_x(domain = c(-4, 4))
```

```{r}
data(seattle)

obsplot(seattle) |>
    mark_ruleY(0) |>
    mark_ruleX(x = "date", y1 = "temp_min", y2 = "temp_max", stroke = "temp_max") |>
    scale_color(scheme = "BuRd") |>
    scale_y(
        grid = TRUE, label = "↑ Temperature (°F)",
        transform = JS("d => d * 9 / 5 + 32")
    )
```

```{r}
data(aapl)
aapl120 <- tail(aapl, 120)

obsplot(aapl120) |>
    mark_ruleX(x = "Date", strokeOpacity = 0.1) |>
    mark_ruleX(x = "Date", y1 = "Low", y2 = "High") |>
    mark_ruleX(
        x = "Date", y1 = "Open", y2 = "Close",
        stroke = JS("d => Math.sign(d.Close - d.Open)"),
        strokeWidth = 4,
        strokeLinecap = "round"
    ) |>
    scale_x(label = NULL) |>
    scale_y(grid = TRUE, label = "↑ Stock price ($)") |>
    scale_color(range = c("#e41a1c", "#000000", "#4daf4a")) |>
    opts(inset = 6)

```

```{r}
data(simpsons)

obsplot(simpsons) |>
    mark_ruleX(
        transform_groupX(
            list(y1 = "min", y2 = "max"),
            x = "season", y = "imdb_rating"
        )
    ) |>
    mark_line(
        transform_groupX(
            y = "median",
            x = "season", y = "imdb_rating", stroke = "red"
        )
    ) |>
    mark_dot(x = "season", y = "imdb_rating") |>
    scale_x(type = "point", label = "Season →", labelAnchor = "right") |>
    scale_y(label = "↑ IMDb rating")
```


## Text mark

```{r}
data(alphabet)

obsplot(alphabet) |>
    mark_barY(x = "letter", y = "frequency") |>
    mark_text(
        x = "letter", y = "frequency",
        text = JS("d => (d.frequency * 100).toFixed(1)"), dy = -5
    ) |>
    mark_ruleY(0) |>
    scale_x(label = NULL) |>
    scale_y(grid = TRUE, label = "↑ Frequency (%)", transform = JS("d => d * 100"))
```

In the following example, the `text` channels passed to `mark_text` are not data column names but explicit values. In Observable Plot this is done by setting to an array (`text: ["2019"]`). In `obsplot`, we use the helper function `as_data` to achieve the same goal.

```{r}
data(travelers)
last_traveler <- head(travelers, 1)

obsplot(travelers) |>
    mark_ruleY(0) |>
    mark_line(x = "date", y = "previous", stroke = "#bab0ab") |>
    mark_line(x = "date", y = "current") |>
    mark_text(
        last_traveler, x = "date", y = "previous",
        fill = "#8a817c", text = as_data("2019"), dy = "-0.5em"
    ) |>
    mark_text(
        last_traveler, x = "date", y = "current",
        text = as_data("2020"), dy = "1.2em"
    ) |>
    scale_y(
        grid = TRUE, nice = TRUE, label = "↑ Travelers per day (millions)",
        transform = JS("d => d / 1e6")
    )
```

```{r}
library(stringr)

data(caltrain)
north <- subset(caltrain, orientation == "N")
south <- subset(caltrain, orientation == "S")

obsplot(caltrain, height = 480, width = 240) |>
    mark_text(x = 1, y = 3, text = as_data("Northbound"), textAnchor = "start") |>
    mark_text(x = -1, y = 3, text = as_data("Southbound"), textAnchor = "end") |>
    mark_text(
        unique(caltrain$hours), x = 0, y = JS("d => d"),
        text = JS('d => `${(d - 1) % 12 + 1}${(d % 24) >= 12 ? "p": "a"}`')
    ) |>
    mark_text(
        north,
        transform_stackX2(
            x = 1,
            y = "hours",
            text = str_pad(north$minutes, 2, side = "left", pad = 0),
            title = paste(north$time, north$line),
            fill = "type",
            textAnchor = "start"
        )
    ) |>
    mark_text(
        south,
        transform_stackX2(
            x = -1,
            y = "hours",
            text = str_pad(south$minutes, 2, side = "left", pad = 0),
            title = paste(north$time, north$line),
            fill = "type",
            textAnchor = "end"
        )
    ) |>
    mark_ruleX(c(-.5, .5)) |>
    scale_x(axis = NULL) |>
    scale_y(domain = 3:26, axis = NULL) |>
    scale_color(domain = "NLB", range = c("currentColor", "peru", "brown"))
```



## Tick mark

```{r}
data(stateage)

obsplot(stateage, height = 300) |>
    mark_ruleX(0) |>
    mark_tickX(
        transform_normalizeX(basis = "sum", z = "name", x = "population", y = "age")
    ) |>
    scale_x(axis = "top", label = "Percent (%) →", transform = JS("d => d*100")) |>
    scale_y(domain = unique(stateage$age), label = "Age") |>
    opts(marginLeft = 50, grid = TRUE)
```

```{r}
obsplot(height = 60) |>
    mark_tickX(list(length = 500), x = JS("d3.randomNormal()"), strokeOpacity = 0.2) |>
    scale_x(domain = c(-4, 4))
```

```{r}
data(morley)

obsplot(morley, height = 150) |>
    mark_ruleY(
        transform_groupY(
            list(
                x1 = JS('V => Math.max(d3.min(V), d3.quantile(V, 0.25) * 2.5 - d3.quantile(V, 0.75) * 1.5)'),
                x2 = JS('V => Math.min(d3.max(V), d3.quantile(V, 0.75) * 2.5 - d3.quantile(V, 0.25) * 1.5)')
            ),
            x = "Speed", y = "Expt"
        )
    ) |>
    mark_barX(
        transform_groupY(
            list(
                x1 = JS('V => d3.quantile(V, 0.25)'),
                x2 = JS('V => d3.quantile(V, 0.75)')
            ),
            x = "Speed", y = "Expt", fill = "#ccc"
        )
    ) |>
    mark_tickX(
        transform_groupY(
            list(x = JS('d3.median')),
            x = "Speed", y = "Expt", strokeWidth = 2
        )
    ) |>
    scale_x(grid = TRUE, inset = 6)
```
