---
title: "obsplot Gallery : Plot changelog"
author: "Julien Barnier"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 4.5
    toc: true
vignette: >
  %\VignetteIndexEntry{obsplot Gallery : Plot changelog}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
library(htmlwidgets)
knitr::opts_chunk$set(
    screenshot.force = FALSE,
    echo = TRUE
)
options("obsplot_menu" = FALSE)
library(obsplot)
library(dplyr)
```

This document tries to reproduce charts included in the [Observable Plot Changelog](https://github.com/observablehq/plot/blob/main/CHANGELOG.md).


## Plot 0.2.0

### Render functions

Marks can be passed as render functions. This can be achieved in `obsplot` with `mark_function()` or `mark_svg()` :

```{r}
data(aapl)
obsplot(aapl) |>
  mark_line(x = "Date", y = "Close") |>
  mark_svg("<text x=20% y=20% fill=red>Hello, world!</text>")
```

When using `mark_function`, the function definition must be enclosed into an `htmlwidgets::JS` call. The `svg` template literal has been imported from Observable standard library and can be used directly :

```{r}
obsplot(aapl) |>
  mark_line(x = "Date", y = "Close") |>
  mark_function(JS("() => svg`<text x=20% y=20% fill=red>Hello, world!</text>`"))
```

### Marks

All marks now accept a `shapeRendering` option :

```{r}
obsplot() |>
  mark_dot(x = 0, y = 0, r = 100, shapeRendering = "crispEdges")
```

`strokeWidth` option can be specified as a channel :
```{r}
df <- data.frame(value = 1:40, stroke = (1:40) / 15)
obsplot(df, height = 60) |>
  mark_dotX(x = "value", strokeWidth = "stroke")
```

`stroke` and `strokeOpacity` for `text` mark can now be specified as channels :

```{r}
df <- data.frame(value = LETTERS[1:10], x = 1:10, opacity = seq(0, 1, length.out = 10))
obsplot(df, height = 60) |>
  mark_textX(x = "x", text = "value", stroke = "x", strokeOpacity = "opacity")
```


Collapsed domains handling in marks that represent continuous intervals :

```{r}
obsplot(df) |>
  mark_rectY(data = c(1, 1, 1), transform_binX())
```

### Scales

New `sort` options :

```{r}
data(alphabet)
obsplot(alphabet) |>
  mark_barY(
    x = "letter", y = "frequency",
    sort = list(x = "y", reverse = TRUE)
  )
```

New *threshold* scale type :

```{r}
obsplot(alphabet) |>
  mark_barY(
    x = "letter", y = "frequency", fill = "frequency",
    sort = list(x = "y", reverse = TRUE)
  ) |>
  scale_y(percent = TRUE) |>
  scale_color(type = "threshold", domain = 0.05)
```

New *diverging* color scales options :

```{r}
data(gistemp)
g <- obsplot(gistemp) |>
  mark_dot(x = "Date", y = "Anomaly", stroke = "Anomaly") |>
  mark_ruleY(0) |>
  opts(grid = TRUE)

g |>
  scale_color(
    type = "diverging", scheme = "BuRd"
  )
```

```{r}
g |>
  scale_color(
    type = "diverging", scheme = "BuRd", symmetric = FALSE
  )
```

```{r}
g |>
  scale_color(type = "diverging-sqrt")
```

New axis *line* option :

```{r}
obsplot() |>
  opts(grid = TRUE, inset = 6) |>
  scale_x(domain = "ABCDEFGH") |>
  scale_y(domain = c(0, 1), line = TRUE)
```

### Facets

Empty facets are no longer rendered :

```{r}
library(palmerpenguins)
obsplot(penguins) |>
  mark_dot(x = "bill_depth_mm", y = "bill_length_mm") |>
  mark_frame() |>
  facet(x = "sex", y = "species") |>
  opts(grid = TRUE)
```

### Transforms

New *filter*, *sort* and *reverse* options to bin and group transforms *outputs* objects :

```{r}
data(sfcases)
df <- sfcases |> filter(case_disposition == "Death")
obsplot(df) |>
  mark_areaY(
    transform_binX(
      list(y = "sum", filter = NULL),
      x = "specimen_collection_date",
      y = "case_count",
      fill = "transmission_category",
      curve = "step",
      thresholds = JS("d3.utcWeek")
    )
  ) |>
  mark_ruleY(0)
```

New reducers in *bin* and *group* transforms :

```{r}
obsplot(trends2020, height = 900) |>
  mark_areaY(x = "week", y = "interest", sort = list(fy = "y", reduce = "max-index")) |>
  facet(y = "search", marginLeft = 160) |>
  opts(label = NULL) |>
  scale_x(axis = "top") |>
  scale_y(axis = NULL)
```

*k* can be passed directly as first argument to window transforms :

```{r}
obsplot(gistemp) |>
  mark_ruleY(0) |>
  mark_line(transform_windowY(24, list(x = "Date", y = "Anomaly")))
```
